---
title: "Intro to Transcriptomics"
author: Nicole Black, Wade Boohar, Kayla Xu
date: 07/17/22
updated: 10/18/24
---

***Deliverables***
-Upload this R Notebook to your GitHub and submit the link to your Repo on Brightspace.
-Include any graphs or figures created in this assignment in the folder with your R notebook with descriptive file names.

Since this is an optional partner activity, it is okay if your answers are the same as your partner’s as long as everyone understands it and could explain it in their own words if asked. Each person must individually push their code to Github. *At the top of your R Notebook, write the name of you and your partner(s) as a comment.*

***Complete the following coding activity and answer any following questions as comments in your R Notebook***

In SummarizedExperiment Tutorial, you learned how to manipulate the SummarizedExperiment data structure and turn it into more readable dataframes, saving them as rna_counts, rna_clinical, and rna_genes. In this semi-guided assignment, you will use these dataframes to perform differential expression analysis based on tumor status.

*Pre-Assignment*
Use knitr function to set your working directory to your analysis_data folder in 490_cluster.
```{r setup}
 knitr::opts_knit$set(root.dir = normalizePath("/home1/arsalans/analysis_data"))
```

If DESeq2 is not already installed, install it now
```{r setup}
if (!require("DESeq2", quietly = TRUE))
BiocManager::install("DESeq2")
```

Load in all necessary packages
```{r setup}
if (!require("knitr"
, quietly = TRUE))
install.packages("knitr")
if (!require("BiocManager"
, quietly = TRUE))
install.packages("BiocManager")
BiocManager::install(version = "3.21")
if (!require("TCGAbiolinks"
, quietly = TRUE))
BiocManager::install("TCGAbiolinks")
library(BiocManager)
library(TCGAbiolinks)
BiocManager::install("EnhancedVolcano")
library(EnhancedVolcano)
```



*1*
Read in the rna_clinical, rna_genes, and rna_counts dataframes which you made in the "SummarizedExperiment Guided Tutorial" R Notebook

```{r setup}
rna_clinical <- read.csv("brca_rna_clinical_data.csv", row.names = 1)

rna_genes <- read.csv("brca_rna_gene_data.csv", row.names = NULL)

rna_counts <- read.csv("brca_rna_count_data.csv", row.names = 1)

```


*2*
In this assignment, you will run differential expression analysis comparing patient samples by whether the sample is from a tumor or normal tissue (this is the definition column in rna_clinical). You will need to choose a variable to control for covariance of: age and/or PAM50 subtype (paper_BRCA_Subtype_PAM50). 

Manipulate those columns so that they are ready for differential expression analysis (hint: what kind of variables are they? what data type are they by default? do you need to handle unknown values?) Filter out genes with a total expression across all patients less than 1000.
```{r}
library(DESeq2)

# variable of interest = tumor or normal tissue = shortLetterCode
rna_clinical$shortLetterCode <- factor(rna_clinical$shortLetterCode)

# same factoring for covariables
rna_clinical$age_category <- factor(rna_clinical$age_category)
rna_clinical$paper_BRCA_Subtype_PAM50 <- factor(rna_clinical$paper_BRCA_Subtype_PAM50)

rna_clinical <- rna_clinical[!is.na(rna_clinical$shortLetterCode) &!is.na(rna_clinical$age_category),]
none_va <- !is.na(rna_clinical$shortLetterCode) &!is.na(rna_clinical$age_category)
rna_counts <- rna_counts[none_va,]

x <- as.character(rna_clinical$paper_BRCA_Subtype_PAM50)
x[is.na(x) | x == "[Not Available]"] <- "none"
rna_clinical$paper_BRCA_Subtype_PAM50 <- factor(x)

rna_clinical$paper_BRCA_Subtype_PAM50[apply(rna_clinical == "none", 1, any)]

summary(rna_clinical$paper_BRCA_Subtype_PAM50)
class(rna_clinical)

rna_genes <- rna_genes$gene_id
rna_counts <- rna_genes$gene_id
view(rna_counts)
rna_clinical # finish line


# pre-filtering
keep_genes1 <- rowSums(rna_counts) >= 10
filtered_rna_counts <- rna_counts[keep_genes,]

```


```{r}
#1
# DESeq2 preprocessing
library(DESeq2)

#2
# Setting variables
rna_clinical <- as.data.frame(rna_se@colData) # clinical info
rna_counts <- as.data.frame(rna_se@assays@data$unstranded) # gene count
rna_genes <- as.data.frame(rna_se@rowRanges@elementMetadata) # 
rownames(rna_genes) <- rna_genes$gene_id # match gene_id
rownames(rna_counts) <- rna_genes$gene_id # match gene_id with rna_counts
colnames(rna_counts) <- rownames(rna_clinical) # match rows of rna_clinical with columns of rna_counts

#3
# renaming values for ease of reading
rna_clinical$shortLetterCode[rna_clinical$shortLetterCode=="TP"] <- "Tumor Primary"
rna_clinical$shortLetterCode[rna_clinical$shortLetterCode=="NT"] <- "Normal Tissue"
rna_clinical$shortLetterCode <- factor(rna_clinical$shortLetterCode)

#4
# removing NA values
sLCmask <- !is.na(rna_clinical$shortLetterCode) # this is for removing NA values in sLC
agemask <- !is.na(rna_clinical$age_at_index) # this is for removing NA values in age
rna_clinical <- rna_clinical[sLCmask & agemask, ]

#5
# align samples with each other
samemask  <- colnames(rna_counts) %in% rownames(rna_clinical)
rna_counts <- rna_counts[, samemask] ## this is currently outputting 0 variables which is not right (check #2) ##
rna_clinical <- rna_clinical[colnames(rna_counts), ]

#6
# this is the filter the genes and then realign them w rna_counts
keep_genes2 <- rowSums(rna_counts) >= 1000 # filter genes with >=1000
rna_counts <- rna_counts[keep_genes2, ]


```


*3*
Perform the differential expression analysis, All you need to do is fill in the appropriate # terms
```{r}

# creating DESeq2 Data Set
# creating dds variable
dds <- DESeqDataSetFromMatrix(countData = rna_counts,
                              colData = rna_clinical,
                              design = ~ age_at_index + shortLetterCode)

# Run DESeq2
dds_obj <- DESeq(dds)

resultsNames(dds_obj)

# Get results comparing Tumor vs Normal
results <- results(dds_obj, format = "DataFrame", 
                  contrast = c("shortLetterCode", "Tumor Primary", "Normal Tissue"))

results <- data.frame(results)

```

Prepare results dataframe for EnhancedVolcano plotting. Add two columns, "-log10(padj)" and "gene_name". Fill in these columns appropriately.
```{r}

# preprocessing
padjmask <- !is.na(results$padj)
results <- results[padjmask,]

# labelling
results$gene_name <- rna_genes$gene_name[match(rownames(results), rna_genes$gene_id)]
results$minuslog10apdj <- -log10(padjmask)

```

*4*
Now we will use the EnhancedVolcano package to plot our results. The code is already completed and should run without adjustment if all code up to here is correct.
```{r}
EnhancedVolcano(results,
                lab = results$gene_name,
                x = 'log2FoldChange',
                y = 'minuslog10apdj',
                title = 'Sample Definition: Tumor vs Normal Tissue',
                pointSize = 1.0,
                labSize = 5.0)

library(ggplot2)
ggsave("/home1/arsalans/490_cluster/Week07_Transcriptomics/EVPlot.png")
```

*5*
# Explain what genes from each part of the Volcano Plot mean in terms of their significance and up/down regulation. 
top-right genes: upregulated and significant
bottom-right genes: upregulated but not significant
top-left genes: downregulated and significant
bottom-left genes: downregulated and not significant
top-middle genes: significant but not up or down regulated
bottom-middle genes: not significant nor up or down regulated

Save the picture of the volcano plot (using either ggsave() or right clicking and manually downloading the image and push this .Rmd and the image to GitHub)rna_clinical$shortLetterCode <- factor(rna_clinical$shortLetterCode)